Why there are more species in several small patches vs few large patches:

### 2024-09-26 ms_SLOSS.Rmd
Submited version

While the overall habitat loss is widely accepted as a process leading to long-term biodiversity loss [@Keil2015;@Hodgson2011], the proposition that single large reserves retain more species than several small reserves of the same area (SL > SS) remain at a similar hierarchical state, even after several criticisms (reviewed in  @Fahrig2017). Empirical observational and experimental data shows the predominance of the opposite pattern (SS > SL) ever since the 80s and this fact has been reinforced to this day [@Simberloff1982; @Fahrig2020; @Hammill2020]. The actual processes driving it remain elusive and it is probably multi-causal. In that scenario, model simulations could produce useful hypotheses to better understand the processes leading to non-intuitive patterns of biodiversity loss.

@@

This research aims to investigate multi-species spatially explicit models, incorporating a fixed level of habitat destruction, neutral and competitive community dynamics, diverse dispersal-related traits, and various geometries of habitat loss. The principal objective is to verify the conditions under which the SS > SL pattern holds and discern the underlying causes. Through this examination, we seek to contribute insights that can help determine if the conditions outlined in the SLOSS cube hypothesis are supported by our modeling approach..

@@

In neutral models, where species are considered equivalent, one species, by chance, tends to dominate a patch [@Hubbell2001]. Consequently, in fragmented landscapes, once a species increases its density, the likelihood of another species replacing it is low. With more patches, the probability of having a greater number of species is heightened. This effect is further amplified when individuals can actively select their habitat, leading to increased survival of propagules. Consequently, fragmentation exhibits a positive influence on the number of species in such scenarios.

@@

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also, an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioural characteristics of the species not considered in the model, including environmental heterogeneity, matrix permeability, and individual’s cost of dispersal, among others. Further, the size of the focal landscape compared to dispersal distances, which determines the scale of the dynamics, could influence quantitative results [@Rybicki2020]. Additionally, the application of mean parameter values in relatively simple multispecies models has been successfully employed and tested for predicting species biomass, as demonstrated by @Fort2018.

>>>>

While the overall habitat loss is widely accepted as a process leading to long-term biodiversity loss [@Keil2015;@Hodgson2011], the proposition that single large reserves retain more species than several small reserves of the same area (SL > SS) remain at a similar hierarchical state, even after several criticisms [reviewed in  @Fahrig2017]. Empirical observational and experimental data shows the predominance of the opposite pattern (SS > SL) ever since the 80s and this fact has been reinforced to this day [@Simberloff1982; @Fahrig2020; @Hammill2020]. The actual processes driving it remain elusive and it is probably multi-causal. In that scenario, model simulations could produce useful hypotheses to better understand the processes leading to non-intuitive patterns of biodiversity loss.

@@

This research aims to investigate multi-species spatially explicit models, incorporating a fixed level of habitat destruction, neutral and competitive community dynamics, diverse dispersal-related traits, and various geometries of habitat loss. The principal objective is to verify the conditions under which the SS > SL pattern holds and discern the underlying causes. Through this examination, we seek to contribute insights that can help determine if the conditions outlined in the SLOSS cube hypothesis are supported by our modeling approach.

@@

In neutral models, where species are considered equivalent, one species, by chance, tends to dominate a patch [@Hubbell2001]. Consequently, in fragmented landscapes, once a species increases its density, the likelihood of another species replacing is low. With more patches, the probability of having a greater number of species is heightened. This effect is further amplified when individuals can actively select their habitat, leading to increased survival of propagules. Consequently, fragmentation exhibits a positive influence on the number of species in such scenarios.

@@

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our model to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also, an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioural characteristics of the species not considered in the model, including environmental heterogeneity, matrix permeability, and individual’s cost of dispersal, among others. Further, the size of the focal landscape compared to dispersal distances, which determines the scale of the dynamics, could influence quantitative results [@Rybicki2020]. Additionally, the application of mean parameter values in relatively simple multispecies models has been successfully employed and tested for predicting species biomass, as demonstrated by @Fort2018.
